Spatial population structure determines extinction risk in climate-induced range shifts

### 2019-06-27 Manuscript/AmNat_ReSubmission.tex
Updated the manuscript with the revisions from the editor's comments

\section*{Results}
\subsection*{Extinction}
Extinction dynamics during the climate-driven range shifts varied widely among scenarios, ranging from no extinctions in some scenarios to the extinction of every simulated population in others. We quantified extinction probability in each scenario as the proportion of simulated populations to decline to extinction through time during the $100$ generations of climate change. Comparing extinction probabilities among scenarios revealed four prominent patterns. First, the slope of the environmental gradient strongly impacted extinction risk (Fig. 2). Scenarios with a flat gradient only resulted in extinctions at a fast speed of climate change coupled with a stark range edge (Fig. B2), while scenarios with a steep gradient experienced high extinction probabilities at even a slow speed of climate change (Fig. B1). Second, edge type also had an important effect on extinction probability, with stark edges increasing extinction risk over gradual and moderate edges. Interestingly, when scenarios with moderate and gradual range edges diverged in extinction probabilities, moderate edges consistently yielded lower extinction risk than scenarios with gradual edges (Fig. 2, B1, \& B2). Third, while both the environmental gradient and edge type influenced extinction probabilities, the environmental gradient drove more dramatic changes to extinction risk. We varied both parameters widely among scenarios, doubling the slope of the environmental gradient and increasing edge starkness by a factor of $100$ (Table A2). Thus, our results suggest the environmental gradient is the stronger driver of extinction risk during climate-induced range shifts across a wide region of parameter space and corresponding biological scenarios. Finally, as expected, the speed of climate change also influenced extinction probability with faster climate change corresponding to greater extinction risk (Fig. B1 \& B2).

@@

\subsection*{Adaptation to the environmental gradient} 
Equilibrium values of the niche trait closely matched the environmental gradient in the range core and diverged at the edges (Fig. 5), likely due to a combination of gene swamping from abundant central populations~\citep{kirkpatrick1997evolution} and the loss of genetic diversity due to drift in small, peripheral populations~\citep{polechova2018sky, polechova2015limits}. In agreement with previous results, the starkness of the range edge influenced this pattern, with increased maladaptation of peripheral populations in scenarios with gradual edges~\citep{garcia1997genetic}. Across all scenarios, the clines in the niche trait became much shallower during the range shift (Fig. 5). As a result, the match between the environmental gradient and local trait values substantially deteriorated during climate change, resulting in reduced fitness across most of the range. This reduction in fitness, observed at all speeds of climate change (Fig. B7 \& B8), could partially explain the increased extinction probability observed for scenarios with a steep environmental gradient (Fig. 2).

@@

Previous research has identified multiple factors underlying increased extinction risk during climate change, including niche breadth and range extent~\citep{thuiller2005niche, schwartz2006predicting}. While we did not vary niche breadth in our model, both edge starkness and the severity of the environmental gradient resulted in decreased range extents and greater extinction probabilities, agreeing with previous findings~\citep{schwartz2006predicting}. Our model framework provides new mechanistic insight, however, indicating that adaptation to an environmental gradient and range edge starkness could be key indicators of increased risk during climate-driven range shifts. A survey of the scientific literature found evidence for local adaption in approximately $71\%$ of studies, suggesting a high prevalence of local adaptation in natural populations~\citep{hereford2009quantitative}. While it is difficult to exactly map adaptation to the environmental gradient to empirical measures of local adaptation, the steepest gradient used in our study resulted in the intrinsic formation of stable range boundaries rather than population collapse, which can occur when the environmental gradient is too steep~\citep{kirkpatrick1997evolution, alleaume2006geographical, polechova2015limits, polechova2018sky}. This suggests the parameters used in the scenarios examined here provide reasonable approximations of empirical patterns. Further, a recent meta-analysis of $1400$ bird, mammal, fish, and tree species found no evidence for consistent declines in abundance towards range edges~\citep{dallas2017species}, a pattern matching that imposed by the starkest range edges examined here (Fig. A3). While some of these patterns could represent a publication bias, for example against negative results in studies of local adaptation, they suggest many species may have spatial population structures similar to those imposed by steep environmental gradients or stark range edges in our model. Our results suggest such spatial population structures can dramatically increase extinction risk in the face of climate change, meaning many species may be at greater risk than previously understood.

@@

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/ResultFigures/MainMeanNiche"}
\caption{Spatial distributions of niche trait phenotypes at three different time points during climate change. Solid lines indicate the average phenotype at each $x$ coordinate and dotted lines are the interquartile ranges of patch means. Means and interquartile ranges were calculated across the pooled patch level means along the $y$ dimension of each simulation at each $x$ coordinate. Graphs show the spatial distributions of niche trait phenotypes in scenarios with (a) steep, (b) shallow, and (c) flat environmental gradients. Colors indicate different edge types as shown in the figure legend. Results are shown for the equilibrium distribution ($t = 0$), halfway through climate change ($t = 50$), and the end of climate change ($t = 100$) for a moderate speed of climate change.}
\label{fig:MeanNiche}
\end{figure}

@@

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/ResultFigures/MainFitGenVar"}
\caption{Spatial distributions of genetic variance in the niche trait at three different time points during climate change. Solid lines indicate the average genetic variance of patches at each $x$ coordinate and dotted lines are the interquartile ranges. Means and interquartile ranges were calculated across the pooled patch level variances along the $y$ dimension of each simulation at each $x$ coordinate. Graphs show the spatial distributions of genetic variance in the niche trait in scenarios with (a) steep, (b) shallow, and (c) flat environmental gradients. Colors indicate different edge types as shown in the figure legend. Results are shown for the equilibrium distribution ($t = 0$), halfway through climate change ($t = 50$), and the end of climate change ($t = 100$) for a moderate speed of climate change.}
\label{fig:FitGenVar}
\end{figure}

@@

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/ResultFigures/SlowMeanNiche"}
\caption{Spatial distributions of niche trait phenotypes at three different time points during climate change. Solid lines indicate the average phenotype at each $x$ coordinate and dotted lines are the interquartile ranges of patch means. Means and interquartile ranges were calculated across the pooled patch level means along the $y$ dimension of each simulation at each $x$ coordinate. Graphs show the spatial distributions of niche trait phenotypes in scenarios with (a) steep, (b) shallow, and (c) flat environmental gradients. Colors indicate different edge types as shown in the figure legend. Results are shown for the equilibrium distribution ($t = 0$), halfway through climate change ($t = 50$), and the end of climate change ($t = 100$) for a slow speed of climate change.}
\label{Fig:MeanNicheSlow}
\end{figure}

@@

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/ResultFigures/FastMeanNiche"}
\caption{Spatial distributions of niche trait phenotypes at three different time points during climate change. Solid lines indicate the average phenotype at each $x$ coordinate and dotted lines are the interquartile ranges of patch means. Means and interquartile ranges were calculated across the pooled patch level means along the $y$ dimension of each simulation at each $x$ coordinate. Graphs show the spatial distributions of niche trait phenotypes in scenarios with (a) steep, (b) shallow, and (c) flat environmental gradients. Colors indicate different edge types as shown in the figure legend. Results are shown for the equilibrium distribution ($t = 0$), halfway through climate change ($t = 50$), and the end of climate change ($t = 100$) for a fast speed of climate change.}
\label{Fig:MeanNicheFast}
\end{figure}

@@

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/ResultFigures/SlowFitGenVar"}
\caption{Spatial distributions of genetic variance in the niche trait at three different time points during climate change. Solid lines indicate the average genetic variance of patches at each $x$ coordinate and dotted lines are the interquartile ranges. Means and interquartile ranges were calculated across the pooled patch level variances along the $y$ dimension of each simulation at each $x$ coordinate. Graphs show the spatial distributions of genetic variance in the niche trait in scenarios with (a) steep, (b) shallow, and (c) flat environmental gradients. Colors indicate different edge types as shown in the figure legend. Results are shown for the equilibrium distribution ($t = 0$), halfway through climate change ($t = 50$), and the end of climate change ($t = 100$) for a slow speed of climate change.}
\label{Fig:FitGenVarSlow}
\end{figure}

@@

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/ResultFigures/FastFitGenVar"}
\caption{Spatial distributions of genetic variance in the niche trait at three different time points during climate change. Solid lines indicate the average genetic variance of patches at each $x$ coordinate and dotted lines are the interquartile ranges. Means and interquartile ranges were calculated across the pooled patch level variances along the $y$ dimension of each simulation at each $x$ coordinate. Graphs show the spatial distributions of genetic variance in the niche trait in scenarios with (a) steep, (b) shallow, and (c) flat environmental gradients. Colors indicate different edge types as shown in the figure legend. Results are shown for the equilibrium distribution ($t = 0$), halfway through climate change ($t = 50$), and the end of climate change ($t = 100$) for a fast speed of climate change.}
\label{Fig:FitGenVarFast}
\end{figure}

>>>>

\section*{Results}
\subsection*{Extinction}
Extinction dynamics during the climate-driven range shifts varied widely among scenarios, ranging from no extinctions in some scenarios to the extinction of every simulated population in others. We quantified extinction probability in each scenario as the proportion of simulated populations to decline to extinction through time during the $100$ generations of climate change. Comparing extinction probabilities among scenarios revealed four prominent patterns. First, the slope of the environmental gradient strongly impacted extinction risk (Fig. 2). Scenarios with a flat gradient only resulted in extinctions at a fast speed of climate change coupled with a stark range edge (Fig. B2), while scenarios with a steep gradient experienced high extinction probabilities at even a slow speed of climate change (Fig. B1). Second, edge type also had an important effect on extinction probability, with stark edges increasing extinction risk over gradual and moderate edges. Interestingly, when scenarios with moderate and gradual range edges diverged in extinction probabilities, moderate edges consistently yielded lower extinction risk than scenarios with gradual edges (Fig. 2, B1, \& B2). This is likely due to the different impacts of gradual versus moderate edges on the evolutionary dynamics of both traits as we discuss below. Third, while both the environmental gradient and edge type influenced extinction probabilities, the environmental gradient drove more dramatic changes to extinction risk. We varied both parameters widely among scenarios, doubling the slope of the environmental gradient and increasing edge starkness by a factor of $100$ (Table A2). Thus, our results suggest the environmental gradient is the stronger driver of extinction risk during climate-induced range shifts across a wide region of parameter space and corresponding biological scenarios. Finally, as expected, the speed of climate change also influenced extinction probability with faster climate change corresponding to greater extinction risk (Fig. B1 \& B2).

@@

\subsection*{Adaptation to the environmental gradient} 
In scenarios defined by a flat environmental gradient, patterns in the niche trait displayed no meaningful spatial variation as the trait optimum was constant across the entire range. Thus, we primarily present results on evolution of the niche trait in the presence of an environmental gradient.

Equilibrium values of the niche trait closely matched the environmental gradient in the range core and diverged at the edges (Fig. 5), likely due to a combination of gene swamping from abundant central populations~\citep{kirkpatrick1997evolution} and the loss of genetic diversity due to drift in small, peripheral populations~\citep{polechova2018sky, polechova2015limits}. In agreement with previous results, the starkness of the range edge influenced the degree of maladaptation (i.e. divergence from the phenotypic optimum) in edge populations~\citep{garcia1997genetic}. In scenarios defined by more gradual edges, peripheral populations exhibited increased maladaptation compared to the peripheral populations of scenarios with stark edges. Across all scenarios, the clines in the niche trait became much shallower during the range shift (Fig. 5). As a result, the match between the environmental gradient and local trait values substantially deteriorated during climate change, resulting in reduced fitness across most of the range. This reduction in fitness, observed at all speeds of climate change (Fig. B7 \& B8), could partially explain the increased extinction probability observed for scenarios with a steep environmental gradient (Fig. 2).

@@

Previous research has identified multiple factors underlying increased extinction risk during climate change, including niche breadth and range extent~\citep{thuiller2005niche, schwartz2006predicting}. While we did not vary niche breadth in our model, both edge starkness and the severity of the environmental gradient resulted in decreased range extents and greater extinction probabilities, agreeing with previous findings~\citep{schwartz2006predicting}. Our model framework provides new mechanistic insight, however, indicating that adaptation to an environmental gradient and range edge starkness could be key indicators of increased risk during climate-driven range shifts. A survey of the scientific literature found evidence for local adaption in approximately $71\%$ of studies, suggesting a high prevalence of local adaptation in natural populations~\citep{hereford2009quantitative}. Our results suggest such adaptation to an environmental gradient corresponds to a greater extinction risk during climate driven range shifts (Fig. 2).  Further, a recent meta-analysis of $1400$ bird, mammal, fish, and tree species found no evidence for consistent declines in abundance towards range edges~\citep{dallas2017species}, a pattern matching that imposed by the starkest range edges examined here (Fig. A3). While some of these patterns could represent a publication bias, for example against negative results in studies of local adaptation, they suggest many species may have spatial population structures similar to those imposed by steep environmental gradients or stark range edges in our model. Our results suggest such spatial population structures can dramatically increase extinction risk in the face of climate change, meaning many species may be at greater risk than previously understood.

While our model suggests certain spatial patterns in both the dispersal and niche traits may be responsible for the increased extinction risk faced by populations in certain scenarios, it is not clear how prevalent such patterns may be in natural populations. For example, while many studies have found evidence for local adaptation~\citep{hereford2009quantitative}, it is difficult to exactly map adaptation to an environmental gradient to empirical measures of local adaptation. However, the steepest gradient used in our study resulted in the intrinsic formation of stable range boundaries rather than population collapse, which can occur when the environmental gradient is too steep~\citep{kirkpatrick1997evolution, alleaume2006geographical, polechova2015limits, polechova2018sky}. This suggests the parameters used in the scenarios examined here provide reasonable approximations of empirical patterns. Additionally, $64\%$ of studies investigating spatial patterns in genetic diversity found evidence for a decline in genetic diversity from core to edge populations~\citep{eckert2008genetic}, as predicted by our model. Matching the model's predictions for the dispersal trait to empirical results, however, returns more mixed results. Edge populations of \textit{Abronia umbellata}, a coastal dune plant, exhibit greater dispersal capability than core populations~\citep{darling2008increased}, matching the patterns produced by our model. However, examination of edge populations in four cricket species revealed little difference in dispersal capabilities relative to core populations~\citep{simmons2004changes}, contradicting our model. Overall, though, empirical studies of spatial variation in dispersal capability within a range are relatively sparse~\citep{hargreaves2014evolution}, making empirical validation of this model prediction difficult. However, our results suggest such research could prove highly informative in predicting a species' fate during climate change.

@@

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/ResultFigures/MainMeanNiche"}
\caption{Spatial distributions of niche trait phenotypes at three different time points during climate change. Solid lines indicate the average phenotype at each $x$ coordinate and dotted lines are the interquartile ranges of patch means. Means and interquartile ranges were calculated across the pooled patch level means along the $y$ dimension of each simulation at each $x$ coordinate. Graphs show the spatial distributions of niche trait phenotypes in scenarios with (a) steep and (b) shallow environmental gradients. Colors indicate different edge types as shown in the figure legend. Results are shown for the equilibrium distribution ($t = 0$), halfway through climate change ($t = 50$), and the end of climate change ($t = 100$) for a moderate speed of climate change.}
\label{fig:MeanNiche}
\end{figure}

@@

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/ResultFigures/MainFitGenVar"}
\caption{Spatial distributions of genetic variance in the niche trait at three different time points during climate change. Solid lines indicate the average genetic variance of patches at each $x$ coordinate and dotted lines are the interquartile ranges. Means and interquartile ranges were calculated across the pooled patch level variances along the $y$ dimension of each simulation at each $x$ coordinate. Graphs show the spatial distributions of niche trait phenotypes in scenarios with (a) steep and (b) shallow environmental gradients. Colors indicate different edge types as shown in the figure legend. Results are shown for the equilibrium distribution ($t = 0$), halfway through climate change ($t = 50$), and the end of climate change ($t = 100$) for a moderate speed of climate change.}
\label{fig:FitGenVar}
\end{figure}

@@

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/ResultFigures/SlowMeanNiche"}
\caption{Spatial distributions of niche trait phenotypes at three different time points during climate change. Solid lines indicate the average phenotype at each $x$ coordinate and dotted lines are the interquartile ranges of patch means. Means and interquartile ranges were calculated across the pooled patch level means along the $y$ dimension of each simulation at each $x$ coordinate. Graphs show the spatial distributions of niche trait phenotypes in scenarios with (a) steep and (b) shallow environmental gradients. Colors indicate different edge types as shown in the figure legend. Results are shown for the equilibrium distribution ($t = 0$), halfway through climate change ($t = 50$), and the end of climate change ($t = 100$) for a slow speed of climate change.}
\label{Fig:MeanNicheSlow}
\end{figure}

@@

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/ResultFigures/FastMeanNiche"}
\caption{Spatial distributions of niche trait phenotypes at three different time points during climate change. Solid lines indicate the average phenotype at each $x$ coordinate and dotted lines are the interquartile ranges of patch means. Means and interquartile ranges were calculated across the pooled patch level means along the $y$ dimension of each simulation at each $x$ coordinate. Graphs show the spatial distributions of niche trait phenotypes in scenarios with (a) steep and (b) shallow environmental gradients. Colors indicate different edge types as shown in the figure legend. Results are shown for the equilibrium distribution ($t = 0$), halfway through climate change ($t = 50$), and the end of climate change ($t = 100$) for a fast speed of climate change.}
\label{Fig:MeanNicheFast}
\end{figure}

@@

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/ResultFigures/SlowFitGenVar"}
\caption{Spatial distributions of genetic variance in the niche trait at three different time points during climate change. Solid lines indicate the average genetic variance of patches at each $x$ coordinate and dotted lines are the interquartile ranges. Means and interquartile ranges were calculated across the pooled patch level variances along the $y$ dimension of each simulation at each $x$ coordinate. Graphs show the spatial distributions of niche trait phenotypes in scenarios with (a) steep and (b) shallow environmental gradients. Colors indicate different edge types as shown in the figure legend. Results are shown for the equilibrium distribution ($t = 0$), halfway through climate change ($t = 50$), and the end of climate change ($t = 100$) for a slow speed of climate change.}
\label{Fig:FitGenVarSlow}
\end{figure}

@@

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/ResultFigures/FastFitGenVar"}
\caption{Spatial distributions of genetic variance in the niche trait at three different time points during climate change. Solid lines indicate the average genetic variance of patches at each $x$ coordinate and dotted lines are the interquartile ranges. Means and interquartile ranges were calculated across the pooled patch level variances along the $y$ dimension of each simulation at each $x$ coordinate. Graphs show the spatial distributions of niche trait phenotypes in scenarios with (a) steep and (b) shallow environmental gradients. Colors indicate different edge types as shown in the figure legend. Results are shown for the equilibrium distribution ($t = 0$), halfway through climate change ($t = 50$), and the end of climate change ($t = 100$) for a fast speed of climate change.}
\label{Fig:FitGenVarFast}
\end{figure}
